---
title: "Homework 3"
author: "Rachael Latimer, Makayla Whitney"
date: "12/1/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Part 1: Predicting a Categorical Outcome using Regularized Logistic Regression
For this assignment you will work with the twitter dataset and try to predict the sentiment of a given tweet. I uploaded the datasets on the website for convenience and to make sure everyone works on the same dataset. Download the tweet dataset using the R code below. Then, also create the blueprint as shown below.

```{r setup}
# Load the following packages needed for modeling in this assignment
  
  require(caret)
  require(recipes)
  require(ranger)

# Import the tweet dataset with embeddings

tweet <- read.csv('https://raw.githubusercontent.com/uo-datasci-specialization/c4-ml-fall-2021/main/content/post/hw2/data/hw1_tweet_final.csv',header=TRUE)

# Recipe for the tweet dataset

blueprint_tweet <- recipe(x  = tweet,
                          vars  = colnames(tweet),
                          roles = c('outcome',rep('predictor',772))) %>%
  step_dummy('month',one_hot=TRUE) %>% 
  step_harmonic('day',frequency=1,cycle_size=7, role='predictor') %>%
  step_harmonic('date',frequency=1,cycle_size=31,role='predictor') %>%
  step_harmonic('hour',frequency=1,cycle_size=24,role='predictor') %>%
  step_normalize(paste0('Dim',1:768)) %>%
  step_normalize(c('day_sin_1','day_cos_1',
                   'date_sin_1','date_cos_1',
                   'hour_sin_1','hour_cos_1')) %>%
  step_rm(c('day','date','hour')) %>%
  step_num2factor(sentiment,
                  transform = function(x) x + 1,
                  levels=c('Negative','Positive'))
```

##Task 1.1. 
Split the original data into two subsets: training and test. Let the training data have the 80% of cases and the test data have the 20% of the cases.
```{r 1.1}


```

##Task 1.2. Use the caret::train() function and ranger engine to train a model with 10-fold cross-validation for predicting the probability of sentiment being positive using a Bagged Trees model with 500 trees.
```{r 1.2}


```

##Task 1.3. 
Use the caret::train() function and ranger engine to train a model with 10-fold cross-validation for predicting the probability of sentiment being positive using a Random Forest model with 500 trees. Set the number of predictors to consider to 250 for each tree while growing a random forest.
```{r 1.3}


```

##Task 1.4 
Evaluate the performance of the Bagged Tree models (1.2) and Random Forest Model (1.3) on the test dataset. Calculate and report logLoss (LL), area under the reciever operating characteristic curver (AUC), overall accuracy (ACC), true positive rate (TPR), true negative rate (TNR), and precision (PRE) for three models. When calculating ACC, TPR, TNR, and PRE, assume that we use a cut-off value of 0.5 for the predicted probabilities. Summarize these numbers in a table like the following. Decide and comment on which model you would use to predict sentiment of a tweet moving forward.
```{r 1.4}


```
		
##Task 1.5 
Compare the performance of the Bagged Trees Model and Random Forest Model in this assignment to the performance of logistic regression models from the previous assignment. Comment on what you observe. Did Bagged Trees or Random Forest Models perform better than Logistic Regression Models?
```{r 1.5}


```

#Part 2: Predicting a Continous Outcome using Regularized Linear Regression
For this assignment you will work with the Oregon dataset and try to predict the scores.I uploaded the datasets on the website for convenience and to make sure everyone works on the same dataset. Download the Oregon testing dataset using the R code below. Then, also create the blueprint as shown below.

```{r Part 2 setup}
# Load the following packages needed for modeling in this assignment
  
  require(caret)
  require(recipes)
  require(ranger)

# Import the oregon dataset

oregon <- read.csv('https://raw.githubusercontent.com/uo-datasci-specialization/c4-ml-fall-2021/main/content/post/hw2/data/hw1_oregon_final.csv',header=TRUE)

# Recipe for the oregon dataset

  outcome <- 'score'
  
  id      <- 'id'

  categorical <- c('sex','ethnic_cd','tst_bnch','migrant_ed_fg','ind_ed_fg',
                   'sp_ed_fg','tag_ed_fg','econ_dsvntg','stay_in_dist',
                   'stay_in_schl','dist_sped','trgt_assist_fg',
                   'ayp_dist_partic','ayp_schl_partic','ayp_dist_prfrm',
                   'ayp_schl_prfrm','rc_dist_partic','rc_schl_partic',
                   'rc_dist_prfrm','rc_schl_prfrm','grp_rpt_dist_partic',
                   'grp_rpt_schl_partic','grp_rpt_dist_prfrm',
                   'grp_rpt_schl_prfrm')

  numeric <- c('enrl_grd')

  cyclic <- c('date','month')


blueprint_oregon <- recipe(x     = oregon,
                    vars  = c(outcome,categorical,numeric,cyclic),
                    roles = c('outcome',rep('predictor',27))) %>%
  step_indicate_na(all_of(categorical),all_of(numeric)) %>%
  step_zv(all_numeric()) %>%
  step_impute_mean(all_of(numeric)) %>%
  step_impute_mode(all_of(categorical)) %>%
  step_harmonic('date',frequency=1,cycle_size=31,role='predictor') %>%
  step_harmonic('month',frequency=1,cycle_size=12,role='predictor') %>%
  step_ns('enrl_grd',deg_free=3) %>%
  step_normalize(c(paste0(numeric,'_ns_1'),paste0(numeric,'_ns_2'),paste0(numeric,'_ns_3'))) %>%
  step_normalize(c("date_sin_1","date_cos_1","month_sin_1","month_cos_1")) %>%
  step_dummy(all_of(categorical),one_hot=TRUE) %>%
  step_rm(c('date','month'))
    
  View(blueprint_oregon %>% prep() %>% summary)

```

##Task 2.1. 
Split the original data into two subsets: training and test. Let the training data have the 80% of cases and the test data have the 20% of the cases.
```{r 2.1}

set.seed(12012021)  # for reproducibility
  
loc      <- sample(1:nrow(oregon), round(nrow(oregon) * 0.8))
ordata_train  <- oregon[loc, ]
ordata_test  <- oregon[-loc, ]

```

##Task 2.2. 
Use the caret::train() function and ranger engine to train a model with 10-fold cross-validation for predicting the scores using a Bagged Trees model with 500 trees.
```{r 2.2}

cv   <- trainControl(method = "none")

grid <- expand.grid(mtry = 27,splitrule='variance',min.node.size=2)
grid

or_baggedtrees <- caret::train(blueprint_oregon,
                             data      = ordata_train,
                             method    = 'ranger',
                             trControl = cv,
                             tuneGrid  = grid,
                             metric = 'logLoss',
                             num.trees = 500,
                             max.depth = 60)
 
or_baggedtrees$times

```

##Task 2.3. 
Use the caret::train() function and ranger engine to train a model with 10-fold cross-validation for predicting the scores using a Random Forest model with 500 trees. Set the number of predictors to consider to 250 for each tree while growing a random forest.
```{r 2.3}
cv   <- trainControl(method = "none")

rgrid <- expand.grid(mtry = 250,splitrule='variance',min.node.size=2)
rgrid

or_rforest <- caret::train(blueprint_oregon,
                        data      = ordata_train,
                        method    = 'ranger',
                        trControl = cv,
                        tuneGrid  = grid,
                        num.trees = 500,
                        max.depth = 60)
 
or_rforest$times

```

##Task 2.4 
Evaluate the performance of the Bagged Tree models (2.2) and Random Forest Model (2.3) on the test dataset. Calculate and report the root mean squared error (RMSE), mean absolute error (MAE), and R-square. 
```{r 2.4}
#checking bagged tree model on the test dataset
ordata_predict_test <- predict(or_baggedtrees, ordata_test)

#calculate RMSE for bagged tree model
bag_rmse <- sqrt(mean((ordata_test$target - ordata_predict_test)^2))
bag_rmse
#for some reason, this is coming up NaN

#calculate MAE for bagged tree model
orpredicted_te <- predict(or_baggedtrees[[500]], ordata_test)

#MAE
bag_mae <- mean(abs(ordata_test$target - orpredicted_te))
bag_mae

#calculate rsqd for bagged tree model
bag_rsqd <- cor(ordata_test$target,orpredicted_te)^2
bag_rsqd



#checking random forest model on the test dataset
rpredict_te <- predict(or_rforest,ordata_test)

# RMSE for random forest model
ran_rmse <- sqrt(mean((ordata_test$target - rpredict_te)^2))
ran_rmse

#calculate MAE for random tree model
test_predict <- predict(or_baggedtrees[[500]], ordata_test)

#MAE
ran_mae <- mean(abs(ordata_test$target - test_predict))
bag_mae

#calculate rsqd for bagged tree model
ran_rsqd <- cor(ordata_test$target, test_predict)^2
bag_rsqd


```

##Task 2.4.2
Summarize these numbers in a table like the following.
```{r 2.4.2}
bagmod <- data.frame(Model = c("Bagged Trees Model"),
                    RMSE = c(bag_rmse),
                     MAE = c(bag_mae),
                     Rsq = c(bag_rsqd))

ranmod <- data.frame(Model = c("Random Forest Model"),
                    RMSE = c(ran_rmse),
                     MAE = c(ran_mae),
                     Rsq = c(ran_rsqd))

#Final Table
SumTable <- rbind(bagmod, ranmod)
SumTable

```

##Task 2.5 
Compare the performance of the Bagged Trees Model and Random Forest Model in this assignment to the performance of linear regression models from the previous assignment. Comment on what you observe. Did Bagged Trees or Random Forest Models perform better than Linear Regression Models in predicting the test scores?
```{r 1.5}



```

